---
title: "Data integration"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/10-data-integration.Rmd/",
  cache.lazy = FALSE,
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r}
pkg <- c("scater", "scran", "uwot", "Rtsne", "batchelor", "EnsDb.Mmusculus.v79", "org.Mm.eg.db", "msigdbr")

lib <- lapply(pkg, library, character.only = TRUE)
```

## Data

### LPMP

Read processed experiment data:

```{r}
sce <- readRDS("data/09-cell-annotation.rds")
```

Make cell names unique across experiments:

```{r}
colnames(sce) <- paste0("lpmp-ehf-", seq_len(ncol(sce)))
```

Make feature names compatible across experiments:

```{r}
rownames(sce) <- rowData(sce)$gene_id
```

Make altExp compatible across experiments:

```{r}
altExp(sce, "ERCC") <- NULL
```

### Gastrulation

Load the MouseGastrulationData package:

```{r}
library(MouseGastrulationData)
```

Select 11c samples:

```{r}
AtlasSampleMetadata <- subset(AtlasSampleMetadata, stage %in% c("E7.25", "E7.5", "E7.75", "E8.0"))
```

Obtain processed samples from the dataset:

```{r}
ref <- EmbryoAtlasData(type = "processed", samples = AtlasSampleMetadata$sample)
```

Filter cells that are labelled as stripped nuclei:

```{r}
ref <- ref[, !ref$stripped]
```

Filter cells that are labelled as doublets:

```{r}
ref <- ref[, !ref$doublet]
```

Filter cells with no celltype label:

```{r}
ref <- ref[, !is.na(ref$celltype)]
```

Compute log-transformed normalized expression values:

```{r}
ref <- logNormCounts(ref)
```

Make cell names unique across experiments:

```{r}
colnames(ref) <- paste0("atlas-", seq_len(ncol(ref)))
```

Make feature names compatible across experiments:

```{r}
rownames(ref) <- rowData(ref)$ENSEMBL
```

Make dimension names compatible across experiments:

```{r}
reducedDimNames(ref) <- c("PCA", "UMAP")
```

## Integration

### Variance modelling

Collect variance modelling statistics:

```{r}
dec.ref <- modelGeneVar(ref, block = ref$sample)

dec.sce <- metadata(sce)$modelGeneVarWithSpikes
```

Make features names compatible across experiments:

```{r}
rownames(dec.ref) <- rowData(ref)$ENSEMBL

rownames(dec.sce) <- rowData(sce)$gene_id
```

Identify common feature names across experiments:

```{r}
ids.ref <- rownames(dec.ref)

ids.sce <- rownames(dec.sce)

ids.all <- intersect(ids.ref, ids.sce)
```

Combine variance modelling results from both studies:

```{r}
dec.ref <- dec.ref[ids.all, ]

dec.sce <- dec.sce[ids.all, ]

dec.all <- combineVar(dec.ref, dec.sce)
```

### Feature selection

### Whitelist genes

Create a whitelist of genes with which to perform feature selection:

```{r}
dec.all$whitelist <- TRUE
```

Remove Xist gene from whitelist:

```{r}
idx <- which(rownames(dec.all) == "ENSMUSG00000086503")

dec.all$whitelist[idx] <- FALSE
```

Remove Y chromosome genes from whitelist:

```{r}
ens <- keys(EnsDb.Mmusculus.v79, keytype = "GENEID", filter = SeqNameFilter("Y"))

idx <- which(rownames(dec.all) %in% ens)

dec.all$whitelist[idx] <- FALSE
```

Remove MT chromosome genes from whitelist:

```{r}
ens <- keys(EnsDb.Mmusculus.v79, keytype = "GENEID", filter = SeqNameFilter("MT"))

idx <- which(rownames(dec.all) %in% ens)

dec.all$whitelist[idx] <- FALSE
```

Remove cell-cycle genes from whitelist:

```{r}
org <- select(org.Mm.eg.db, keys = "GO:0007049", keytype = "GOALL", column = "ENSEMBL")

idx <- which(rownames(dec.all) %in% org$ENSEMBL)

dec.all$whitelist[idx] <- FALSE
```

Remove ribosomal genes from whitelist:

```{r}
sig <- msigdbr(species = "Mus musculus", category = "C2")

sig <- subset(sig, gs_name == "KEGG_RIBOSOME")

org <- select(org.Mm.eg.db, keys = sig$gene_symbol, keytype = "SYMBOL", column = "ENSEMBL")

idx <- which(rownames(dec.all) %in% org$ENSEMBL)

dec.all$whitelist[idx] <- FALSE
```

### Feature selection

Identify common feature names across experiments:

```{r}
ids.ref <- rownames(ref)

ids.sce <- rownames(sce)

ids.all <- intersect(ids.ref, ids.sce)
```

Subset experiments to common universe of features:

```{r}
ref <- ref[ids.all, ]

sce <- sce[ids.all, ]
```

Select highly variable features to use in fastMNN correction:

```{r}
set.seed(294252498)

dec.all <- subset(dec.all, whitelist == TRUE)

hvg <- getTopHVGs(dec.all, var.field = "bio", var.threshold = 0)
```

### Batch normalization

Define source label before splitting experiments:

```{r}
ref$source <- "atlas"

sce$source <- "lpmp-ehf"
```

Define batch label before splitting experiments:

```{r}
ref$batch <- paste0("atlas-", ref$sample)

sce$batch <- "lpmp-ehf"
```

Define cluster label before splitting experiments:

```{r}
ref$cluster <- paste0("atlas-", ref$cluster)

sce$cluster <- paste0("lpmp-ehf-", sce$cluster)
```

Split experiments by batch into a list:

```{r}
fct <- unique(ref$sample)

all <- lapply(fct, function(x) ref[, ref$sample == x])

all <- c(all, sce)

lvl <- paste0("atlas-", fct)

names(all) <- c(lvl, "lpmp-ehf")
```

Perform batch normalization:

```{r}
all <- do.call(multiBatchNorm, all)
```

### Batch integration

Clear incompatible data from all experiments before merging:

```{r}
all <- lapply(all, function(x) { metadata(x) <- list(); x })

all <- lapply(all, function(x) { assays(x) <- assays(x)[c("counts", "logcounts")]; x })

all <- lapply(all, function(x) { rowData(x) <- NULL; x })

var <- c("source", "batch", "cluster", "celltype", "stage")

all <- lapply(all, function(x) { colData(x) <- colData(x)[, var]; x })

all <- lapply(all, function(x) { reducedDims(x) <- list(); x })

rng <- rowRanges(all[[1]])

all <- lapply(all, function(x) { rowRanges(x) <- rng; x })
```

Create combined experiment object for integration:

```{r}
all <- do.call(cbind, all)
```

Create merge order for experiments:

```{r}
lvl <- c("E6.5", "E6.75", "E7.0", "mixed_gastrulation", "E7.25", "E7.5", "E7.75", "E8.0", "E8.25", "E8.5")

dat <- AtlasSampleMetadata

dat$stage <- factor(dat$stage, levels = lvl)

idx <- order(dat$stage, dat$ncells, decreasing = TRUE)

dat <- dat[idx, ]
  
idx <- as.character(dat$sample)

idx <- paste0("atlas-", idx)

idx <- c(idx, "lpmp-ehf")
```

Correct experiments in order specified above:

```{r}
set.seed(596542508)

mnn <- correctExperiments(
  all,
  batch = all$batch,
  subset.row = hvg,
  correct.all = TRUE,
  PARAM = FastMnnParam(merge.order = idx)
)
```

Inspect the percentage of variance lost as a diagnostic:

```{r}
metadata(mnn)$merge.info$lost.var
```

## Exploration

Create UMAP matrix using settings from MouseGastrulationData analysis:

```{r}
set.seed(626393619)

mnn <- runUMAP(mnn, dimred = "corrected", n_neighbors = 20, min_dist = 0.7)
```

Create UMAP plot coloured by celltype and facetted by batch:

```{r fig.width = 12, fig.height = 9}
dim <- reducedDim(mnn, "UMAP")

colnames(dim) <- c("X", "Y")

dat <- as.data.frame(mnn@colData)

tbl <- cbind(dim, dat)

tbl$batch <- factor(tbl$batch, levels = idx)

tbl$stage <- "atlas"

tbl$stage[tbl$batch == "lpmp-ehf"] <- "lpmp-ehf"

ggplot(tbl, aes(X, Y, colour = celltype)) + 
  geom_point(data = tbl[, c("X", "Y")], colour = "#F0F0F0") + 
  geom_point(size = 0.5) + 
  scale_colour_manual(values = MouseGastrulationData::EmbryoCelltypeColours) + 
  guides(colour = guide_legend(override.aes = list(size = 1.5))) + 
  facet_wrap(~ stage) + 
  theme_void() + 
  theme(
    aspect.ratio = 1,
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

## Summary

### Ouptut

Save integrated experiment object to disk:

```{r}
saveRDS(mnn, file = "data/10-data-integration.rds")
```

### Session

Print session information:

```{r}
sessionInfo()
```
